MS&E 226: Project Part 1 Code

### 2024-11-29 MS&E 226_Project Part 2_Code.Rmd
started part 2rmd file

# Split the data into features and target variable
X <- df_clean_scaled[, !names(df_clean) %in% "Uses.Consumed.Binary"] # continuous outcome variable and linear combination of other columns
# Step 1: Prepare the data
# Identify numeric and categorical columns
numeric_cols <- sapply(X, is.numeric)
categorical_cols <- sapply(X, function(x) is.factor(x) | is.character(x))

# Create dummy variables for categorical features
dummy <- dummyVars(" ~ .", data = X[, categorical_cols])
categorical_dummy <- predict(dummy, newdata = X[, categorical_cols])

# Combine numeric and dummy-coded categorical variables
X <- cbind(X[, numeric_cols], categorical_dummy)

@@

# Step 4: Train the final k-NN model with the best k
final_model <- knn(train = X, test = X, cl = y, k = best_k[[1]])

@@

@@

# Train the logistic regression model with cross-validation and lambda tuning
model_cv <- train(
  formula,
  data = df_clean_scaled,
  method = "glmnet",
  family = "binomial",
  trControl = ctrl,
  metric = "Accuracy",
  tuneGrid = tune_grid
)

@@

# Train final model on all data with the best lambda
final_model <- glmnet(
  as.matrix(df_clean_scaled[, all.vars(formula)[-1]]), 
  df_clean_scaled$Uses.Consumed.Binary, 
  alpha = 0,
  lambda = model_cv$bestTune$lambda,
  family = "binomial"
)

@@

# Get the best model
best_model <- model_grid$finalModel

@@

>>>>

# Split the data into features and target variable
X <- df_clean_scaled[, !names(df_clean) %in% "Uses.Consumed.Binary"] # continuous outcome variable and linear combination of other columns
# Step 1: Prepare the data
# Identify numeric and categorical columns
knn_numeric_cols <- sapply(X, is.numeric)
knn_categorical_cols <- sapply(X, function(x) is.factor(x) | is.character(x))

# Create dummy variables for categorical features
dummy <- dummyVars(" ~ .", data = X[, knn_categorical_cols])
categorical_dummy <- predict(dummy, newdata = X[, knn_categorical_cols])

# Combine numeric and dummy-coded categorical variables
X <- cbind(X[, knn_numeric_cols], categorical_dummy)

@@

# Step 4: Train the final k-NN model with the best k
knn_class_final_model <- knn(train = X, test = X, cl = y, k = best_k[[1]])

@@

# Create model formula
lr_formula <- as.formula("Uses.Consumed.Binary ~ Utility.Type + Max.Revenue.Sector + Demand.Winter.Peak + Demand.Summer.Peak + Revenue.Retail + Revenue.Retail:Max.Revenue.Sector")

@@

# Train the logistic regression model with cross-validation and lambda tuning
model_cv <- train(
  lr_formula,
  data = df_clean_scaled,
  method = "glmnet",
  family = "binomial",
  trControl = ctrl,
  metric = "Accuracy",
  tuneGrid = tune_grid
)

@@

# Train final model on all data with the best lambda
lr_class_final_model <- glmnet(
  as.matrix(df_clean_scaled[, all.vars(lr_formula)[-1]]), 
  df_clean_scaled$Uses.Consumed.Binary, 
  alpha = 0,
  lambda = model_cv$bestTune$lambda,
  family = "binomial"
)

@@

# Get the best model
dt_class_final_model <- model_grid$finalModel

@@

# Project Part 2

## 1: Prediction on the holdout set

### Classification

#### a)

In part 1, the logistic regression classifier had the best performance. This is the classifier that will be used for prediction on the holdout set.

```{r Classification prediction with winning modeling strategy}
# Select specific columns from the dataframe based on positive correlation > 0.1 or negative correlation <-0.1
holdout_clean <- holdout[, c("Uses.Consumed.Binary", # Binary outcome variable
                   "Utility.Type", # Positively or negatively correlated with outcome, depending on factor level 
                   "Demand.Winter.Peak", # Positively correlated with outcome
                   "Demand.Summer.Peak", # Positively correlated with outcome
                   "Max.Revenue.Sector", # Negatively correlated with outcome
                   "Revenue.Retail", # Positively correlated with outcome
                   grep("^Retail\\.", names(df), value = TRUE))] # Positively correlated with outcome

# Loop through each column in the dataset and apply the factor transformation to each non-numeric (categorical) variable
holdout_clean[] <- lapply(holdout_clean, function(x) {
  if (!is.numeric(x)) {
    factor(x, levels = unique(x))
  } else {
    x
  }
})

holdout_clean$Uses.Consumed.Binary <- factor(holdout_clean$Uses.Consumed.Binary, levels = c("0", "1"))
levels(holdout_clean$Uses.Consumed.Binary) <- make.names(levels(holdout_clean$Uses.Consumed.Binary))

# Scale numeric columns (use same scaling parameters as training data)
holdout_clean_scaled <- holdout_clean
numeric_cols_holdout <- sapply(holdout_clean, is.numeric)  # Identify numeric columns
holdout_clean_scaled[numeric_cols_holdout] <- scale(
  holdout_clean[numeric_cols_holdout], 
  center = attr(scale(df_clean[numeric_cols]), "scaled:center"), 
  scale = attr(scale(df_clean[numeric_cols]), "scaled:scale")
)

# Subset the holdout dataset to include only variables in the formula
required_columns <- all.vars(formula)[-1]  # Exclude the response variable
holdout_clean <- holdout[, required_columns, drop = FALSE]  # Keep only required columns

# subset only covariates in the formula 
formula_vars <- all.vars(lr_formula)

# Subset the dataframe to include only the columns in the formula
holdout_clean_scaled <- holdout_clean_scaled[, formula_vars, drop = FALSE]

# Separate outcome and covarites

Y_holdout <- holdout_clean_scaled$Uses.Consumed.Binary

X_holdout <- holdout_clean_scaled[, colnames(holdout_clean_scaled) != "Uses.Consumed.Binary"]

# Convert to matrix for glmnet predictions
holdout_class_matrix <- as.matrix(X_holdout)

# Make predictions (probabilities)
predictions_prob <- predict(
  lr_class_final_model, 
  newx = holdout_class_matrix, 
  s = model_cv$bestTune$lambda, 
  type = "response"
)

# Convert probabilities to binary predictions
predictions_binary <- ifelse(predictions_prob >= 0.5, "1", "0")

# Calculate 0-1 loss
true_labels <- holdout$Uses.Consumed.Binary  # True labels from holdout
misclassified <- sum(predictions_binary != true_labels)
zero_one_loss_holdout <- misclassified / length(true_labels)

# Print results
print(paste("0-1 Loss on holdout data:", round(zero_one_loss_holdout, 4)))

## 2: Inference

#### (c)

```{r fit model on the holdout set}

# Define the columns to keep 
column_names <- c(
  "Demand.Summer.Peak", "Demand.Winter.Peak", "Sources.Generation", "Sources.Purchased",
  "Sources.Other", "Uses.Retail", "Uses.Resale", "Uses.No.Charge", "Uses.Consumed", "Uses.Losses",
  "Retail.Residential.Customers", "Retail.Commercial.Customers", "Retail.Industrial.Customers",
  "Retail.Transportation.Customers", "Utility.Region", "Max.Revenue.Sector", "Revenue.Total", "Utility.Type"
)

# Create holdout_pred with only the specified columns from holdout
holdout_pred <- holdout[, column_names]

### Applying data transformations
# Applying log transformation to the Revenue.Total column
holdout_pred$Log_Revenue_Total <- log1p(holdout_pred$Revenue.Total)

# Prepare the holdout dataset
holdout_pred <- holdout[, column_names]
holdout_pred$Log_Revenue_Total <- log(holdout_pred$Revenue.Total)

# Prepare the data
X <- model.matrix(Log_Revenue_Total ~ ., data = holdout_pred)[, -1]  # Design matrix (exclude intercept)
y <- holdout_pred$Log_Revenue_Total  # Response variable

# Prepare the input matrices for glmnet
x <- model.matrix(Log_Revenue_Total ~ ., data = df_pred)[, -1]  # Remove intercept column
y <- df_pred$Log_Revenue_Total

# Fit the Ridge regression model with predefined lambda
ridge_model <- glmnet(x, y, alpha = 0, lambda = final_lambda_ridge, standardize = TRUE)

# View model coefficients
print(coef(ridge_model))
